Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-04-12 paper.tex
update
small changes

1. The galactic centre (g.c.). This is where all numerical simulations
of cold dark matter predict the highest density. However, the detailed
dark matter density in the very central part is difficult to predict,
due a to a possibly very complicated interplay between baryons, dark
matter, and the central galactic black hole. Also, it is a very
crowded region with many gamma-ray sources like pulsars and other
supernova recants, which have to be subtracted from the data to
extract the dark matter induced signal. In fact, there is a recent
claim of an indication of a relatively light dark matter particle
contribution to the gamma-ray flux from the
g.c. \cite{2010arXiv1010.2752H}, but other hypotheses seem to work at
least as well \cite{2010arXiv1012.5839B}.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v10.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v10.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v10.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v10.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that inverse
  Compton upscattered both CMB photons (orange dash-dotted) as well as
  dust and star photons (green dotted), and CRp induced gamma-ray
  emission (red solid). Each panel is associated with an individual DM
  BM model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$ and a boost from
  substructures of 570. We find bright prospects for detecting the
  continuum emission that is dominating the total emission in the GeV
  energy range. Also note that, above 100~MeV the total inverse
  Compton emission is at least a factor 10 smaller than the emission
  expected from both the continuum emission and the emission coming
  from CRps.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.v10.0.1deg.1.6T.SubMass.elmu.SF300.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.BMv10.0.1deg.SubMass.noMW.woGal.eps}
\caption{\it Comparing the flux from different inverse Compton
  upscattered radiation fields. We show the differential inverse
  Compton emission induced by leptophilic DM in the left panel and by
  the $\Kp$ benchmark model in the right panel. The contribution from
  each individual radiation field from top line to bottom line; CMB
  (green dashed-double-dotted), dust (blue dashed), stars (purple
  dashed-dotted). The sum of the three components are shown with the
  orange solid line. The black dotted line represents the upscattered
  CMB photons without any cooling from stars and dust. The red solid
  line show the CRp induced gamma-ray flux. The black arrow show the
  differential upper limit from Fermi \cite{2010ApJ...717L..71A}. All
  fluxes are calculated for the Fornax cluster within $\rvir$ using a
  point spread function of $0.1\deg$. The boost from Sommerfeld and
  substructures is about 130 and 570, respectively.}
 \label{fig:IR_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v10.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v10.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v10.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v10.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Comparing the emission from different clusters. We show
  the differential flux from four clusters; Fornax (upper left), Coma
  (upper right), Virgo (lower left), and Perseus (lower right). The
  different flux models; CRp induced emission (red solid), leptophilic
  models (LP) include both final state radiation and IC upscattered
  CMB, dust and starlight (green dash-dotted), and benchmark $\Kp$
  models (BM) include continuum emission, and IC upscattered CMB, dust
  and starlight (blue dashed). The arrows have colors matching each
  cluster and show the differential upper limits set by Fermi-LAT in
  the energy ranges $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from
  left to right. For visualization we offset the arrows by increasing
  the mean energy by 30\% in the opposite order as they appear in the
  figure, i.e. starting with Perseus. The flux from the clusters is
  integrated out to $\rvir$ using a point spread function of $0.1\deg$
  and includes the boost from substructures. We find that the lower
  GeV energy regime is most constraining, and induce upper limits on
  boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/LP.const.diff.v10.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Constraining boost factors using flux upper limits. The
  leptophilic (LP) gamma-ray emission is derived within $\rvir$ using
  a point spread function of $0.1\deg$. In order not to overproduce
  the Fermi-LAT differential flux upper limits in the energy interval
  $1-10$~GeV the boost from substructures and Sommerfeld enhancement
  (SFE) is constrained for four clusters; Fornax (red), Virgo
  (orange), M49 (green), and NGC4636 (Blue). The SFE has been rescaled
  from 300 (refer to equation) that is required to explain the
  electron and positron excess observed at Earth with LP DM (black
  arrow). If the DM interpretation is correct, we can constrain the
  smallest size of halos to be larger than $10\,\msun$. However, if
  the smallest size of DM halos is $10^{-6}\,\msun$, we can constrain
  the SFE to about 4 in Fornax, corresponding to a maximum SFE in the
  MW of about 10.}
 \label{fig:boost_const}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/SB.v10.1GeV.SF300.SubMass.elmu.eps}
\caption{\it Comparing the surface brightness from different
  clusters. We show the CRp induced emission (red), leptophilic
  emission (green), and emission from the $\Kp$ benchmark model
  (blue). The surface brightness above 1~GeV is derived using a very
  small point spread function ($\psf=10^{-5}\deg$). The
  different linestyles each represent a cluster; Fornax (solid), Coma
  (dotted), Virgo (dashed), and Perseus (dash-dotted). The boost from
  Sommerfeld and substructures is about 130 and 570,
  respectively. Notice the very similar shape and normalization of the DM
  models, while the CRps have a much larger scatter in the profiles.}
 \label{fig:SB_clu}
\end{figure}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v10.1GeV.SF300.noSuB.vs.SubMass.elmu.eps}
\caption{\it The impact of substructures on surface brightness. We
  show the surface brightness at 1~GeV for the Fornax cluster using a
  very small point spread function of $10^{-5}\deg$ that reveals the
  details of the spatial profiles. The emission induced by CRps is
  denoted by the red solid line, the leptophilic models by green
  lines, and the DM $\Kp$ benchmark model by blue lines. The dashed
  line show the brightness where the boost from substructures is
  included while the substructures are excluded in the dotted
  line. The boost from Sommerfeld and substructures is about 130 and
  570, respectively. Notice the nearly flat profiles when
  substructures are included, and the relative large boost at a
  percent of $\rvir$.}
 \label{fig:SB_sub}
\end{figure}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{figures/SB.v10.1GeV.SF300.noSuB.elmu.eps}
  \includegraphics[width=0.49\columnwidth]{figures/SB.fornax.v10.1GeV.SF300.noSuB.elmu.eps}
\caption{\it The surface brightness without substructures. We show the
  CRp induced emission (red), leptophilic emission (green), and
  emission from the $\Kp$ benchmark model (blue). The surface
  brightness above 1~GeV is derived using a very small point spread
  function ($\psf=10^{-5}\deg$). LEFT PANEL: compare the
  surface brightness for different clusters; Fornax (solid), Coma
  (dotted), Virgo (dashed), and Perseus (dash-dotted). RIGHT PANEL:
  compare different emission components in Fornax; dotted lines show
  the inverse Compton (IC) upscattered CMB photons, dashed lines show
  the IC upscattered photons from stars and dust, and blue solid line
  show the continuum emission from the $\Kp$ benchmark
  model.}
 \label{fig:SB_clu_nosub}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{figures/SB.Fornax.v10.SF300.SubMass.elmu.eps}
  \includegraphics[width=0.49\columnwidth]{figures/SB.Virgo.v10.SF300.SubMass.elmu.eps}
\caption{\it Surface brightness predicted for Cherenkov telescopes at
  high energies. We show the emission above $100$~GeV and include the
  boost from substructures. We use a point spread function of
  $\psf=0.1\deg$ that is typical for both Cherenkov
  telescopes as well as the Fermi-LAT at this energy. Left panel show
  the Fornax cluster and right panel the Virgo cluster. The gamma-ray
  emission is derived for the following components; CRps (red solid),
  continuum emission from the DM $\Kp$ benchmark model (blue
  dash-dotted), as well as final state radiation (green dashed) and
  inverse Compton upscattered dust and starlight (green dotted) from
  leptophilic DM.}
 \label{fig:SB_IACTs}
\end{minipage}
\end{figure*}

>>>>

1. The galactic centre (g.c.). This is where all numerical simulations
of cold dark matter predict the highest density. However, the detailed
dark matter density in the very central part is difficult to predict,
due a to a possibly very complicated interplay between baryons, dark
matter, and the central galactic black hole. Also, it is a very
crowded region with many gamma-ray sources like pulsars and other
supernova remnants, which have to be subtracted from the data to
extract the dark matter induced signal. In fact, there is a recent
claim of an indication of a relatively light dark matter particle
contribution to the gamma-ray flux from the
g.c. \cite{2010arXiv1010.2752H}, but other hypotheses seem to work at
least as well \cite{2010arXiv1012.5839B}.

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.BMcompI.v9.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJ.v9.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompK.v9.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.BMcompJs.v9.0.1deg.1.6T.SubMass.IR2.noMW.woGal.eps}
\caption{\it Comparing the differential flux from different models: we
  show the continuum emission from DM benchmark (BM) models (blue
  dashed), electrons and positrons from DM BM models that inverse
  Compton upscattered both CMB photons (orange dash-dotted) as well as
  dust and star photons (green dotted), and CRp induced gamma-ray
  emission (red solid). Each panel is associated with an individual DM
  BM model; upper left $\Ip$, lower left $\Kp$, upper right $\Jp$, and
  lower right $\Js$. The emission is calculated for the Fornax cluster
  using a point spread function of $0.1\deg$ and a boost from
  substructures of 570. We find bright prospects for detecting the
  continuum emission that is dominating the total emission in the GeV
  energy range. Also note that, above 100~MeV the total inverse
  Compton emission is at least a factor 10 smaller than the emission
  expected from both the continuum emission and the emission coming
  from CRps.}
 \label{fig:diff_BM}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.v9.0.1deg.1.6T.SubMass.elmu.SF300.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.IRcomp.BMv9.0.1deg.SubMass.noMW.woGal.eps}
\caption{\it Comparing the flux from different inverse Compton
  upscattered radiation fields. We show the differential inverse
  Compton emission induced by leptophilic DM in the left panel and by
  the $\Kp$ benchmark model in the right panel. The contribution from
  each individual radiation field from top line to bottom line; CMB
  (green dashed-double-dotted), dust (blue dashed), stars (purple
  dashed-dotted). The sum of the three components are shown with the
  orange solid line. The black dotted line represents the upscattered
  CMB photons without any cooling from stars and dust. The red solid
  line show the CRp induced gamma-ray flux. The black arrow show the
  differential upper limit from Fermi \cite{2010ApJ...717L..71A}. All
  fluxes are calculated for the Fornax cluster within $\rvir$ using a
  point spread function of $0.1\deg$. The boost from Sommerfeld and
  substructures is about 130 and 570, respectively.}
 \label{fig:IR_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v9.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v9.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v9.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v9.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Comparing the emission from different clusters. We show
  the differential flux from four clusters; Fornax (upper left), Coma
  (upper right), Virgo (lower left), and Perseus (lower right). The
  different flux models; CRp induced emission (red solid), leptophilic
  models (LP) include both final state radiation and IC upscattered
  CMB, dust and starlight (green dash-dotted), and benchmark $\Kp$
  models (BM) include continuum emission, and IC upscattered CMB, dust
  and starlight (blue dashed). The arrows have colors matching each
  cluster and show the differential upper limits set by Fermi-LAT in
  the energy ranges $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from
  left to right. For visualization we offset the arrows by increasing
  the mean energy by 30\% in the opposite order as they appear in the
  figure, i.e. starting with Perseus. The flux from the clusters is
  integrated out to $\rvir$ using a point spread function of $0.1\deg$
  and includes the boost from substructures. We find that the lower
  GeV energy regime is most constraining, and induce upper limits on
  boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/LP.const.diff.v9.0.1deg.1.6T.SubMass.SF300.IR2.noMW.woGal.eps}
\caption{\it Constraining boost factors using flux upper limits. The
  leptophilic (LP) gamma-ray emission is derived within $\rvir$ using
  a point spread function of $0.1\deg$. In order not to overproduce
  the Fermi-LAT differential flux upper limits in the energy interval
  $1-10$~GeV the boost from substructures and Sommerfeld enhancement
  (SFE) is constrained for four clusters; Fornax (red), Virgo
  (orange), M49 (green), and NGC4636 (Blue). The SFE has been rescaled
  from 300 (refer to equation) that is required to explain the
  electron and positron excess observed at Earth with LP DM (black
  arrow). If the DM interpretation is correct, we can constrain the
  smallest size of halos to be larger than $10\,\msun$. However, if
  the smallest size of DM halos is $10^{-6}\,\msun$, we can constrain
  the SFE to about 4 in Fornax, corresponding to a maximum SFE in the
  MW of about 10.}
 \label{fig:boost_const}
\end{figure}

@@

\begin{figure}
 \includegraphics[width=0.99\columnwidth]{figures/SB.v9.1GeV.SF300.SubMass.elmu.eps}
\caption{\it Comparing the surface brightness from different
  clusters. We show the CRp induced emission (red), leptophilic
  emission (green), and emission from the $\Kp$ benchmark model
  (blue). The surface brightness above 1~GeV is derived using a very
  small point spread function ($\psf=10^{-5}\deg$). The
  different linestyles each represent a cluster; Fornax (solid), Coma
  (dotted), Virgo (dashed), and Perseus (dash-dotted). The boost from
  Sommerfeld and substructures is about 130 and 570,
  respectively. Notice the very similar shape and normalization of the DM
  models, while the CRps have a much larger scatter in the profiles.}
 \label{fig:SB_clu}
\end{figure}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v9.1GeV.SF300.noSuB.vs.SubMass.elmu.eps}
\caption{\it The impact of substructures on surface brightness. We
  show the surface brightness at 1~GeV for the Fornax cluster using a
  very small point spread function of $10^{-5}\deg$ that reveals the
  details of the spatial profiles. The emission induced by CRps is
  denoted by the red solid line, the leptophilic models by green
  lines, and the DM $\Kp$ benchmark model by blue lines. The dashed
  line show the brightness where the boost from substructures is
  included while the substructures are excluded in the dotted
  line. The boost from Sommerfeld and substructures is about 130 and
  570, respectively. Notice the nearly flat profiles when
  substructures are included, and the relative large boost at a
  percent of $\rvir$.}
 \label{fig:SB_sub}
\end{figure}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{figures/SB.v9.1GeV.SF300.noSuB.elmu.eps}
  \includegraphics[width=0.49\columnwidth]{figures/SB.fornax.v9.1GeV.SF300.noSuB.elmu.eps}
\caption{\it The surface brightness without substructures. We show the
  CRp induced emission (red), leptophilic emission (green), and
  emission from the $\Kp$ benchmark model (blue). The surface
  brightness above 1~GeV is derived using a very small point spread
  function ($\psf=10^{-5}\deg$). LEFT PANEL: compare the
  surface brightness for different clusters; Fornax (solid), Coma
  (dotted), Virgo (dashed), and Perseus (dash-dotted). RIGHT PANEL:
  compare different emission components in Fornax; dotted lines show
  the inverse Compton (IC) upscattered CMB photons, dashed lines show
  the IC upscattered photons from stars and dust, and blue solid line
  show the continuum emission from the $\Kp$ benchmark
  model.}
 \label{fig:SB_clu_nosub}
\end{minipage}
\end{figure*}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.49\columnwidth]{figures/SB.Fornax.v9.SF300.SubMass.elmu.eps}
  \includegraphics[width=0.49\columnwidth]{figures/SB.Virgo.v9.SF300.SubMass.elmu.eps}
\caption{\it Surface brightness predicted for Cherenkov telescopes at
  high energies. We show the emission above $100$~GeV and include the
  boost from substructures. We use a point spread function of
  $\psf=0.1\deg$ that is typical for both Cherenkov
  telescopes as well as the Fermi-LAT at this energy. Left panel show
  the Fornax cluster and right panel the Virgo cluster. The gamma-ray
  emission is derived for the following components; CRps (red solid),
  continuum emission from the DM $\Kp$ benchmark model (blue
  dash-dotted), as well as final state radiation (green dashed) and
  inverse Compton upscattered dust and starlight (green dotted) from
  leptophilic DM.}
 \label{fig:SB_IACTs}
\end{minipage}
\end{figure*}
